---
title: "GPs with Functional Inputs for WEPP"
author: "Jarad Niemi"
date: "2023-06-15"
format: revealjs
---

```{r}
library("tidyverse")
library("knitr")
```

## (Some) Computer Models in Agricultre

```{r}
models <- tribble(
  ~name, ~developer, ~date, 
  "WEPP", "USDA", 1990, 
  "Agro-IBIS", "Wisconsin", 1990,
  "APSIM", "CSIRO (Australia)", 2007, 
  # "APSIM Next Gen", "CSIRO (Australia)", 2014, 
  # "CLEM",NA,NA,"livestock",
  # "IMPACT",NA,NA,NA,
  # "SEAMLESS",NA,NA,NA,
  # "NUANCES-FARMSIM",NA,NA,NA,
  # "FarmDESIGN",NA,NA,NA,
  "Cycles", "Penn State University", 2020, 
  "CyclesL", "Penn State University", 2023,
  "HydroGeoSphere", "Iowa State University", 2023
)

knitr::kable(models)
```

# [Daily Erosion Project](https://www.dailyerosion.org/)

## Daily Erosion Project - Precipitation

![](fig/dep_precipitation)

## Daily Erosion Project - Soil Loss

![](fig/dep_soilloss)

## Daily Erosion Project - Expansion

Currently

- ~200,000 runs/day
- start at midnight, done by 6am
- single server parallelism

Expansion

- Midwest
- U.S.
- World


# [Water Erosion Prediction Project (WEPP)](https://www.fs.usda.gov/ccrc/tool/watershed-erosion-prediction-project-wepp)

## WEPP

![](fig/wepp)

## WEPP Experiment - Profile

![](fig/profile)

## WEPP Experiment - Length and Slope

![](fig/length_and_slope)

## Scientific questions

- Out-of-sample prediction accuracy
- Relative importance
    - Profile
        - Position
    - Length
    - Mean Slope


# Gaussian Process Emulators

## Emulator

Deterministic computer model

$$Y = f(X)$$

for

- input $X$
- output $Y$

An *emulator* is an estimate of $f$, i.e. $\widehat{f}$.

## Gaussian Process

For $f:$ &Xscr; $\to$ &Yscr;, assume
$$f \sim GP(\mu, k)$$

for 

- mean function $\mu:$ &Xscr; $\to \mathbb{R}$ and
- covariance function $k:$ &Xscr; $\times$ &Xscr; $\to \mathbb{R}^{+}$.

For simplicity, $\mu(x) = 0$. 


## Data

For a collection of data $(Y_i, X_i)$ for $i=1,\ldots,n$, 
we have 

$$Y = (Y_1,\ldots,Y_n)^\top \sim N(0, \Sigma)$$

where

$$\Sigma_{i,j} = k(x_i,x_j).$$

For prediction at a location $\widetilde{x}$, 
we have the conditional distribution $\widetilde{y}|Y$
which involves covariances $k(x_i, \widetilde{x})$ for all $i$. 

## Distance-based covariance kernel

For any &Xscr;, let 
$$k(x_i,x_j) = \sigma^2 e^{-d(x_i,x_j)/2}$$
for spatial variance $\sigma^2$ and some distance function $d(x_i,x_j)$. 

For example,

- squared exponential (Gaussian) covariance
- automatic relevance determination
- automatic dynamic relevance determination


## Squared-exponential covariance

If $x \in \mathbb{R}$, the *squared-exponential (Gaussian) covariance* kernel is 
$$d(x_i,x_j) = w (x_i-x_j)^2$$
where $w$ is the weight ($1/w$ is the length-scale/range).


## Automatic relevance determination

If $x \in \mathbb{R}^P$, the *automatic relevance determination* kernel is 
$$d(x_i,x_j) = \sum_{p=1}^P w_p [x_{i,p}-x_{j,p}]^2$$
where $w_p$ controls the strength of the relationship in the $p$th dimension:

- small $w_p$ indicates large differences lead to small distances
- large $w_p$ indicates small differences lead to large distances


## Automatic dynamic relevance determination

If $x \in$ &Hscr; (Hilbert space), 
the *automatic dynamic relevance determination* kernel is 
$$d(x_i,x_j) = \int w(t) [x_{i}(t)-x_{j}(t)]^2 dt$$
where for some weight function $w:$ &Tscr; $\to \mathbb{R}^+$. 
WLOG &Tscr; = [0,1].


# Automatic dynamic relevance determination (ADRD)

## Asymmetric double exponential

![](fig/alf)

## Fourier expansion

![](fig/few)

## B-splines and hinges

![](fig/bsplines)


## Combined

Combine hillslope profile, length, and mean slope into a single correlation
function:

![](fig/combined)

Calculate **scaled-integrated weight**:

![](fig/siw)



# Results

## Out-of-sample prediction accuracy

![](fig/prediction_accuracy)

## Profile, length, and slope relevance

![](fig/estimated_relevance)

## Hillslope profile relevance

![](fig/estimated_weight_function)


# Summary

## Novelty

Introduced

- Automatic dynamic relevance determination
- Variety of weight functions

Results

- Similar prediction accuracy
- Informative relevance


## Future questions

- Calibration
- Curse of big data
- Design
- Shift-invariant distance function
- Data Fusion
    - Field data
    - Remote sensing data
    
## More information

- Slides: <>
- 
